---
title: "class9 halloween project"
author: "Chelsea A16871799"
format: pdf
---

```{r}
candy_file <- read.csv("candy-data.csv") 
candy_file
```

```{r}
candy <- read.csv("candy-data.csv", row.names=1)
head(candy)
```

#Q1. How many different candy types are in this dataset?
```{r}
nrow(candy)
```


#Q2. How many fruity candy types are in the dataset?
```{r}
sum(candy$fruity)
```
Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
```{r}
candy["Almond Joy",]$winpercent
```


Q4. What is the winpercent value for “Kit Kat”?
```{r}
candy["Kit Kat",]$winpercent
```

Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```
```{r}
library("skimr")
skim(candy)
```
Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
#winpercent variable looks to be on different scale to majority of other columns.
Q7. What do you think a zero and one represent for the candy$chocolate column?
#for chocolate column, zero kind of represent false(not chocolate) and 1 represent true(is chocolate candy)

Q8. Plot a histogram of winpercent values
```{r}
hist(candy$winpercent, breaks=30)
```

```{r}
library(ggplot2)
ggplot(candy)+
  aes(winpercent)+
  geom_histogram(binwidth=10)
```

#Q9. Is the distribution of winpercent values symmetrical?
#This distribution of winpercent value is asymmetric.

#Q10. Is the center of the distribution above or below 50%?
The center of distribution is below 50%.

Q11. On average is chocolate candy higher or lower ranked than fruit candy?
```{r}
choc.inds <- as.logical(candy$chocolate)
choc.win <- candy[choc.inds,"winpercent"]
choc.win
```
```{r}
fruit.inds <- as.logical(candy$fruity)
fruit.win <- candy[fruit.inds,"winpercent"]
fruit.win
```

```{r}
mean(choc.win)
mean(fruit.win)
```

On average chocolate candy is higher ranked than fruit candy.
Q12. Is this difference statistically significant?
```{r}
t.test(choc.win,fruit.win)
```
This difference is statistically significant as p value=2.871e-08.

Q13. What are the five least liked candy types in this set?
```{r}
head(candy[order(candy$winpercent),], n=5)
```
Q14. What are the top 5 all time favorite candy types out of this set?
```{r}
head(candy[order(candy$winpercent,decreasing = TRUE ),], n=5)
```

Q15. Make a first barplot of candy ranking based on winpercent values.
```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col(fill="gray")
```
Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```
```{r}
library(ggplot2)
ggplot(candy)+
aes(winpercent, reorder(rownames(candy),winpercent))+
  geom_col(fill=my_cols)
```

- Q17. What is the worst ranked chocolate candy?
Sixlets is the worst ranked chocolate candy.
- Q18. What is the best ranked fruity candy?
starburst is the best ranked fruity candy.

```{r}
library(ggrepel)
#How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 10)
```
Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
Reese's Minature is the the highest ranked for winpercent for least money.

Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
top 5 most expensive candy is Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, Hershey's Milk Chocolate. among this the least popular is the Nik L Nip.
```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```
Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
#chocolaty and fruity are anti-correlated
Q23. Similarly, what two variables are most positively correlated?
#winpercent and chocolate are most positive correlated.
```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
```{r}
plot(pca$x[,1],pca$x[,2])
```
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```
```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
```{r}
library(plotly)
```
```{r}
#ggplotly(p)
```
```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
Fruity, hard, and pluribus.(positive direction)




