---
title: "Class 5 data Visualization with ggplot2"
author: "Chelsea (PID: A16871799)"
format: pdf
---

## Using GGPLOT

The ggplot2 package needs to be installed as it does not come with R " out of the box"

We use the 'install.packages()' to do this

```{r}
head(cars)
```

To use ggplot we first need to load it up before I can call any of functions in the packages. do this with the `library()` function.
```{r}
library(ggplot2)
ggplot()
```


All ggplot find at least 3 thing
-data (stuff we plot)
-aesthetic mapping (aes value)
-geoms

```{r}
ggplot(cars)+ 
  aes(x=speed, y=dist)+
  geom_point()+
  geom_smooth(method="lm", se=FALSE)+
  labs(title="Speed and stopping distance of Cars",
       x="Speed (MPH)", 
       y="Stopping Distance (ft)",
       subtitle = "Your informative subtitle text here",
       caption="Dataset: 'cars'")+
       theme_bw()
  
```
ggplot is not the onlygraphing system in R there are more. There is even "base R" graphics.

```{r}
plot(cars)
```

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
```{r}
nrow(genes)
ncol(genes)
sum(genes$State=="up")
table(genes$State)
round( table(genes$State)/nrow(genes) * 100, 2 )
```


```{r}
p <- ggplot(genes)+
  aes(x=Condition1, y=Condition2, col=State)+
  geom_point()
  
p+scale_colour_manual(values=c("blue","gray","red"))+
  labs(title="Gene Expression Changes Upon Drug Treatment",
       x="Control (no drug)", 
       y="Drug Treatment")
```

# File location online

```{r}
library("dplyr")
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"
gapminder <- read.delim(url)
gapminder_2007 <- gapminder %>% filter(year==2007)
```
```{r}
gapminder_1957 <- gapminder %>% filter(year==1957| year==2007)

ggplot(gapminder_1957) + 
  geom_point(aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop), alpha=0.7) + 
  scale_size_area(max_size = 10) +
  facet_wrap(~year)
```
```{r}
gapminder_top5 <- gapminder %>% 
  filter(year==2007) %>% 
  arrange(desc(pop)) %>% 
  top_n(5, pop)

gapminder_top5
```
```{r}
ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = pop, fill = lifeExp))
```
```{r}
ggplot(gapminder_top5) +
  aes(x=reorder(country, -pop), y=pop, fill=country) +
  geom_col(col="gray30") +
  guides(fill="none")
```





